---
title: "Project 1"
author: "Project 1 Group -- Michelle, Valentina, Ian"
date: "Due date: March 23 2025 at midnight (EST)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this project, you will produce an exploratory data analysis (EDA) report for the `NHANES::NHANESraw` dataset. **Warning:** You are not allowed to use the `NHANES::NHANES` dataset. 

```{r}
install.packages("NHANES")
library(NHANES)
library(ggplot2)
library(dplyr)
NHANESraw
```

## Instruction

1. Define a set of questions/tasks that will guide your exploration. 

# Question 1: Is there an age of peak perceived health? 
```{r}
df <- NHANESraw

# make AgeDecade var
df <- df |>
  mutate(AgeDecade = floor(Age / 10) * 10)

df |>
  select(HealthGen, Gender, AgeDecade) |>
  filter(AgeDecade > 0, !is.na(HealthGen)) |>
  ggplot(aes(x = HealthGen, y = after_stat(prop), group = 1)) + 
  geom_bar() +
  facet_grid(AgeDecade ~ Gender) + 
  labs(
    x="Proportion of Population", 
    y="Self-Reported General Health", 
    title="Proportion of Self-Reported Physical Health by Age Decade and Gender"
  )
```

```{r}
df |>
  filter(!is.na(HealthGen)) |>
  group_by(Age) |>
  summarise(
    prop_good = mean(HealthGen %in% c("Good", "Vgood", "Excellent"))
  ) |>
  ggplot(aes(x = Age, y = prop_good)) +
  geom_line(color = "blue", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  #scale_x_log10() +
  labs(
    x = "Age (years)",
    y = "Proportion with Good/Vgood/Excellent Health",
    title = "Proportion of People Reporting Good, Very Good, or Excellent Health by Age"
  ) +
  theme_minimal()
```

# Question 2: How does pregnancy affect sleep?
i) If a person is pregnant now, how is their sleep different from those who are 
not pregnant now?
```{r}
df <- NHANESraw

df |> 
  select(PregnantNow, SleepHrsNight, Gender) |> 
  filter(PregnantNow %in% c("Yes", "No"), Gender == "female") |> 
  ggplot(aes(x=PregnantNow)) + 
  geom_violin(aes(y=SleepHrsNight)) + 
  labs(
    x="Pregnant", 
    y="Reported Sleep Per Night (Hours)", 
    title="Sleep Quality in Pregnant and Non Pregnant People"
  ) + 
  theme_minimal()
```
ii) Are a larger amount of pregnancies total associated with worse sleep? 
```{r}
df |> 
  select(nPregnancies, Gender, SleepHrsNight,SleepTrouble) |> 
  filter(Gender == "female", nPregnancies <= 20) |> 
  ggplot(aes(x=nPregnancies, y=SleepHrsNight, color=SleepTrouble)) + 
  geom_jitter(width = 0.4, height = 0.4, alpha = 0.4) + 
  scale_x_continuous(breaks = seq(0, 20, by = 2)) + 
  geom_smooth(method = "loess", se=FALSE) + 
  labs(
    x="Number of Pregnancies Had", 
    y="Sleep Per Night (Hours)", 
    title="Sleep Quality vs. Number Pregnancies", 
    color="Sleep Trouble Reported"
  ) + 
  theme_minimal()

```

# Question 3: How is blood pressure associated with age?
```{r}
df_summary <- df |>
  group_by(Age) |>
  summarise(
    med_sys = median(BPSysAve, na.rm = TRUE),
    med_dia = median(BPDiaAve, na.rm = TRUE)
  )

ggplot(df_summary, aes(x = Age)) +
  geom_smooth(aes(y = med_sys, color = "Systolic"), se = FALSE) +
  geom_smooth(aes(y = med_dia, color = "Diastolic"), se = FALSE) +
  scale_color_manual(
    values = c("Systolic" = "orange",
               "Diastolic" = "purple")
  ) +
  labs(
    y = "Median Blood Pressure", 
    x = "Age", 
    title = "Median Blood Pressure By Age", 
    color = "Type"
  ) +
  theme_minimal()
```
# Question 4: How is blood pressure related with BMI?
```{r}
df |> 
  filter(BMI_WHO != "NA") |>
  ggplot(aes(x=BMI_WHO, y=BPSysAve)) +
  geom_violin(fill = "lightblue") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.2) +
  labs(title = "Systolic Blood Pressure Across BMI Categories",
       x = "BMI Category",
       y = "Systolic BP")

df |> 
  filter(BMI_WHO != "NA") |>
  ggplot(aes(x=BMI_WHO, y=BPDiaAve)) +
  geom_violin(fill = "lightblue") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.2) +
  labs(title = "Systolic Blood Pressure Across BMI Categories",
       x = "BMI Category",
       y = "Diastolic BP")
```
# Question 5: Do people with high incomes typically have better mental health?
```{r}
df <- NHANESraw |>
  select(Age, HHIncome, DaysMentHlthBad, Depressed) |>
  filter(Age >= 18) |>
  filter(!is.na(HHIncome)) |>
  filter(!is.na(DaysMentHlthBad)) |>
  filter(!is.na(Depressed))

summary_df <- df |>
  group_by(HHIncome) |>
  summarise(avg_bad_days = mean(DaysMentHlthBad))
```

```{r}
ggplot(summary_df, aes(x = HHIncome, y = avg_bad_days, fill = avg_bad_days)) +
  geom_col() +
  scale_fill_gradient(low = "tomato", high = "steelblue",
                      name = "Avg. Bad Days") +
  labs(
    title   = "Mental Health by Household Income",
    x       = "Household Income",
    y       = "Avg. Bad Mental Health Days (per month)",
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(df, aes(x = HHIncome, y = DaysMentHlthBad)) +
  geom_boxplot(fill = "lightblue", color = "steelblue") +
  labs(
    title    = "Spread of Bad Mental Health Days by Income",
    x        = "Household Income",
    y        = "Bad Mental Health Days (per month)",
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

2. Create 6-10 figures from the data exploring your questions.
  
    - at least one figure must be faceted using `facet_`
    - each counted figure must be multivariate and show at least two of the plotting techniques we learned in class
    - all figures must be well-labeled
    - no more than two figures of the same type (counted using `geom_`, if you include $>8$ figures, you are allowed to have more than 2 of a given type)

3. Write a report with your figures included. 

## Submission

- A `zip` file consisting of one `RMD` and one `HTML` files. The name of your `zip` file should be a two-digit number corresponding to your group. 
- The knitted output of the `RMD` file must be identical to the `HTML` file.
- The `HTML` file should only consist of words and plots, i.e., no code. To do this, you can set `echo = F` in the global options.  

## Grading structure

The precise rubric will be published via ELMS. You will be graded on:

  1. Basic completion according to the instruction
  2. Depths of your guiding questions
  3. Suitable plot types
  4. Meaningful statistical analysis/transforms to support your questions and provide additional insights
  5. Statistical rigor, including but not restricted to: expressing uncertainty, explanation why/how the methods work, reproducibility of the experiments
  
## Group Structure 

The following applies to a standard 3-person group. I will make suitable adjustment for the non-standard group. 

| Role | Responsibilities |
|------|----------------|
| **Project Leader/Manager** | - Guides the overall direction of the EDA project <br> - Organizes workflow, deadlines, communication (with team members and the instructor), and file submission  |
|||
| **Data/Stat Lead** | - Performs data cleaning, transformations, and statistical analysis<br>- Provides technical support to others |
|||
| **Visualization/Communication Lead** | - Designs clear, well-labeled, and statistically informed figures<br>- Standardizes figures/texts and integrate them into report<br>|
|||
| **All Members** | - Each member is responsible for at $\geq 2$ plots, and must draft texts for their own plots <br> - Each member is responsible for writing a post-production report |

> **Note:** Roles have some degree of flexiblility; members can overlap responsibilities. Every member must contribute technically, write text, and support quality control.

## Post-production

Each member is required to submit a report including an individual reflection and a group member evaluation. Both will be available only to the instructor/grader. Each report must include the following key information:

- Your roles and your plots (1 paragraph)
- Summary of your work (1 paragraph)
- Evaluation of each of your team member's work, i.e., what aspects are done well and what could be improved for each member. (1 paragraph for each member)